Parametricity Consequences of GADTs Representations

### 2021-02-23 20-gadts/GADT-story.tex
work on gadt-story

These kinds of fixpoint equations are entirely sensible at the level
of types. But to ensure that the syntactic fixpoint representing an
ADT actually denotes a semantic object computed as a real-live
fixpoint, the semantic fixpoint calculation must converge.  If we
interpret our types as sets, then this means that the fixpoint being
taken must be of a {\em functor} on the category $\mathit{Set}$ of
sets and functions between them, rather than of a mere function
between sets~\cite{tfca}.  That is, the function $F$ interpreting the
type constructor $\mathsf{F}$ constructing the body of a syntactic
fixpoint must not only have an action on sets, but must also have a
{\em functorial action} on functions between sets. Reflecting this
requirement back into syntax gives that $\mathsf{F}$ must support a
suitable\footnote{The $\mathsf{map}$ function for $\mathsf{F}$ should
  satisfy syntactic reflections of the functoriality properties ---
  i.e., should preserve identities and compositions --- in
  $\mathit{Set}$ even though there is no mechanism in Haskell for
  enforcing this.} $\mathsf{map}$ function, i.e., must be an instance
of Haskell's $\mathsf{Functor}$ class.

Requiring $F$ to be a functor will ensure that the interpretation $\mu
F$ of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure
that $\mu F$ is itself a functor, so that the type constructor
$\mathsf{T}$ associated with the ADT $\mathsf{T\,a}$ supports a
$\mathsf{map}$ function as well, $F$ must again be the fixpoint $\mu
H$ of a functor $H$. In this case, however, $H$ must be a functor on
the category $\mathit{Func(Set)}$ of functors and natural
transformations on $\mathit{Set}$, i.e., $H$ must be a {\em
  higher-order} functor on $\mathit{Set}$. Reflecting this requirement
back into syntax gives that $\mathsf{T\,a} = \mathsf{(\mu F)\,a} =
\mathsf{(\mu\,(\mu H))\,a}$ for the ``type constructor constructor''
$\mathsf{H}$ that supports suitable\footnote{{\color{blue} The
    $\mathsf{map}$ function for $\mathsf{H}$ should satisfy syntactic
    reflections of the functoriality properties in
    $\mathit{Func(Set)}$ --- i.e., should preserve identity natural
    transformations and composition of natural transformations --- and
    the $\mathsf{map}$ function for $\mathsf{H\,F}$ should satisfy
    syntactic reflections of the functoriality properties in
    $\mathit{Set}$ as in the previous footnote. Again, there is no
    mechanism in Haskell for enforcing this.}} $\mathsf{map}$
functions.  {\color{blue} Mention naturality?}

@@

A concrete example is given by the ADT $\mathsf{List\,a}$. This type
is is modelled as the fixpoint $\mu F_{\mathit{List\,a}}$ of the
first-order functor whose action on sets is given by
$F_{\mathit{List}\,a}\,X = 1 + a \times X$ and whose action on
functions is given by $F_{\mathit{List}\,a}\,f = 1 + \id_a \times
f$. The type constructor $\mathsf{List}$ is modelled by the
higher-order functor $H$ whose action on a functor $F$ is given by the
functor $H\,F$ whose actions on sets and functions between them are
given by $H\,F\,X = 1 + X \times F\,X$ and $H\,F\,f = \id_1 + f \times
F\,f$, respectively, and whose action on a natural transformation
$\eta$ is the natural transformation given by $H\,\eta = \id_1 + \eta
\times F\,\eta$. The type constructor $\mathsf{List}$ is then
modelled as the functor $\mu H$. Reflecting the functorial action of
$\mu H$ back into syntax gives Haskell's built-in $\mathsf{map}$
function as the type-and-data-uniform, structure-preserving,
data-changing function associated with $\mathsf{List}$. {\color{blue}
  HOW?}

In a parametric model, short cut fusion can be used to show that the
Church encoding and the primitive representation of an ADT are
isomorphic as sets. Note, however, that whereas it is impossible even
to {\em state} the rearrange-transform property for Church encodings
of ADTs, it is simply the reflection back into syntax of the
{\color{blue} naturality property - right?} of the functorial action
of the interpretation of the type constructor associated with an ADT's
primitive representation.

@@

@@

GADTs relax the restriction on the type instances appearing in a data
type definition. The special form of GADTs known as {\em nested types}
allow the data constructors of a GADT to take as arguments data whose
types involve type instances of the GADT other than the one being
defined. For example, the definition
\[\mathsf{PTree \, a = PLeaf \,a \,|\, PNode \,(PTree\,(a \times
  a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees
introduces the data constructor(s $\mathsf{PLeaf :: a \to PTree\,a}$
and) $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
not only the invariant that all of the data in a structure of type
$\mathsf{PTree\,a}$ is of the same type $\mathsf{a}$, but also the
invariant that all perfect trees have lengths that are powers of
2. More general GADTs allow their constructors both to take as
arguments {\em and return as results} data whose types involve type
instances of the GADT other than the one being defined. An example is
the following GADT, given in Haskell syntax:
\[\begin{array}{l}
\mathsf{data\, Seq :: * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}\]
Since the return type of the data constructor $\mathsf{Pair}$ is not
simply a variable, $\mathsf{Seq}$ is not a nested type.

By contrast with the ADT $\mathsf{List\, a}$, where the type parameter
$\mathsf{a}$ is integral to the type being defined, the type parameter
$\mathsf{a}$ in $\mathsf{PTree\,a}$ and $\mathsf{Seq\,a}$ appears as a
``dummy'' parameter used only to give the kind \verb|* -> *| of the
type constructors $\mathsf{PTree}$ and $\mathsf{Seq}$. This is
explicitly captured by the above GADT syntax. A GADT --- even a nested
type --- thus does not define a {\em family of inductive types}, one
for each type argument, like an ADT does, but instead defines an
entire family of types that must be constructed simultaneously. That
is, a GADT defines an {\em inductive family of types}.

>>>>

These kinds of fixpoint equations are entirely sensible at the level
of types. But to ensure that the syntactic fixpoint representing an
ADT actually denotes a semantic object computed as a real-live
fixpoint, the semantic fixpoint calculation must converge.  If we
interpret our types as sets, then this means that the fixpoint being
taken must be of a {\em functor} on the category $\mathit{Set}$ of
sets and functions between them, rather than of a mere function
between sets~\cite{tfca}.  That is, the function $F$ interpreting the
type constructor $\mathsf{F}$ constructing the body of a syntactic
fixpoint must not only have an action on sets, but must also have a
{\em functorial action} on functions between sets. Reflecting this
requirement back into syntax gives that $\mathsf{F}$ must support a
suitable\footnote{The $\mathsf{map}$ function for $\mathsf{F}$ should
  satisfy syntactic reflections of the functoriality properties in
  $\mathit{Set}$ --- i.e., preservation of identity functions and
  composition of functions --- even though there is no mechanism in
  Haskell for enforcing this.} $\mathsf{map}$ function, i.e., must be
an instance of Haskell's $\mathsf{Functor}$ class.

Requiring $F$ to be a functor will ensure that the interpretation $\mu
F$ of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure
that $\mu F$ is itself a functor, so that the type constructor
$\mathsf{T}$ associated with the ADT $\mathsf{T\,a}$ supports a
$\mathsf{map}$ function as well, we can require that $F$ also be the
fixpoint $\mu H$ of a functor $H$. In this case, however, $H$ must be
a functor on the category $\mathit{Set}^\mathit{Set}$ of functors and
natural transformations on $\mathit{Set}$. That is, $H$ must be a {\em
  higher-order} functor on $\mathit{Set}$. Reflecting this requirement
back into syntax gives that $\mathsf{T\,a} = \mathsf{(\mu F)\,a} =
\mathsf{(\mu\,(\mu H))\,a}$ for the ``type constructor constructor''
$\mathsf{H}$ that supports suitable\footnote{The $\mathsf{map}$
  function for $\mathsf{H}$ should satisfy syntactic reflections of
  the functoriality properties in $\mathit{Set}^\mathit{Set}$ ---
  i.e., preservation of identity natural transformations and
  composition of natural transformations --- and the $\mathsf{map}$
  function for $\mathsf{H\,F}$ should satisfy syntactic reflections of
  the functoriality properties in $\mathit{Set}$ as in the previous
  footnote. Again, there is no mechanism in Haskell for enforcing
  this.} $\mathsf{map}$ functions.

@@

A concrete example is given by the ADT $\mathsf{List\,a}$. This type
is modelled as the fixpoint $\mu F_{\mathit{List\,a}}$ of the
first-order functor whose action on sets is given by
$F_{\mathit{List}\,a}\,X = 1 + a \times X$ and whose action on
functions is given by $F_{\mathit{List}\,a}\,f = 1 + \id_a \times
f$. The type constructor $\mathsf{List}$ is modelled by the functor
that is the fixpoint $\mu H$ of the higher-order functor $H$ whose
action on a functor $F$ is given by the functor $H\,F$ whose actions
on sets and functions between them are given by $H\,F\,X = 1 + X
\times F\,X$ and $H\,F\,f = \id_1 + f \times F\,f$, respectively, and
whose action on a natural transformation $\eta$ is the natural
transformation given by $H\,\eta = \id + \eta \times
F\,\eta$. Reflecting the functorial action of $\mu H$ back into syntax
gives exactly Haskell's built-in $\mathsf{map}$ function as the
type-and-data-uniform, structure-preserving, data-changing function
associated with $\mathsf{List}$. {\color{blue} Double-check this
  paragraph.}

In a parametric model, short cut fusion can be used to show that the
Church encoding and the primitive representation of an ADT are
isomorphic as sets. Note, however, that whereas it is impossible even
to {\em state} the rearrange-transform property for Church encodings
of ADTs, this property is simply the reflection back into syntax of
the instance of naturality for the type-independent function that
rearranges structures of type $\mathsf{T\,a}$ into ones of type
$\mathsf{T'\,a}$ and the functions $\mathsf{map_T\,f}$ and
$\mathsf{map_{T'}\,f}$ for a function $\mathsf{f :: a \to b}$ and the
type constructors $\mathsf{T}$ and $\mathsf{T'}$ associated with these
ADTs, respectively. {\color{blue} Double-check this paragraph.}

@@

GADTs relax the restriction on the type instances appearing in a data
type definition. The special form of GADTs known as {\em nested types}
allow the data constructors of a GADT to take as arguments data whose
types involve type instances of the GADT other than the one being
defined. For example, the definition
\[\mathsf{PTree \, a = PLeaf \,a \,|\, PNode \,(PTree\,(a \times
  a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees
introduces the data constructor(s $\mathsf{PLeaf :: a \to PTree\,a}$
and) $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
not only the invariant that all of the data in a structure of type
$\mathsf{PTree\,a}$ is of the same type $\mathsf{a}$, but also the
invariant that all perfect trees have lengths that are powers of
2. More general GADTs allow their constructors both to take as
arguments {\em and return as results} data whose types involve type
instances of the GADT other than the one being defined. An example is
the GADT $\mathsf{Seq}$ given in~\eqref{eq:seq}. Note that since the
return type of the data constructor $\mathsf{Pair}$ is not of the form
$\mathsf{Seq\,a}$ for any variable $\mathsf{a}$, $\mathsf{Seq}$ is a
GADT that is not a nested type.

By contrast with the ADT $\mathsf{List\, a}$, where the type parameter
$\mathsf{a}$ is integral to the type being defined, the type parameter
$\mathsf{a}$ appears in both $\mathsf{PTree\,a}$ and $\mathsf{Seq\,a}$
as a ``dummy'' parameter used only to give the kind \verb|* -> *| of
the type constructors $\mathsf{PTree}$ and $\mathsf{Seq}$. This is
explicitly captured in the alternative Haskell syntax {\color{blue}
    Correct?}
\[\begin{array}{l}
\mathsf{data\, Seq :: * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}\]
A GADT --- even a nested type --- thus does not define a {\em family
  of inductive types}, one for each type argument, like an ADT does,
but instead defines an entire family of types that must be constructed
simultaneously. That is, a GADT defines an {\em inductive family of
  types}.

@@
